---
title: "Application of the Cophylospace Approach"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Supporting information to
# Next-generation cophylogeny: unravelling eco-evolutionary processes
<font size = 3> Isabel Blasco-Costa<sup>1,2</sup>, Alexander Hayward<sup>3</sup>, Robert Poulin<sup>4</sup> and Juan A. Balbuena<sup>5</sup></font><br>
<font size = 1> <sup>1</sup>Department of Invertebrates, Natural History Museum of Geneva, PO Box 6434, CH-1211 Geneva 6, Switzerland. ORCID: 0000-0002-5001-5986<br>
<sup>2</sup>Department of Arctic and Marine Biology, UiT The Arctic University of Norway, Langnes, PO Box 6050, 9037 Tromsø, Norway.<br>
<sup>3</sup>Centre for Ecology and Conservation, University of Exeter, Penryn Campus, Penryn, Cornwall, Exeter, TR10 9FE, UK. ORCID: 0000-0001-7413-718X<br>
<sup>4</sup>Department of Zoology, University of Otago, PO Box 56, Dunedin New Zealand. ORCID: 0000-0003-1390-1206<br>
<sup>5</sup>Cavanilles Institute of Biodiversity and Evolutionary Biology, University of Valencia, PO Box 22085, 46071 Valencia, Spain. ORCID: 0000-0003-4006-1353<br></font>

We illustrate the application of the Cophylospace approach in R (R Core Team 2021) with an example of 14 *Ligophorus* spp. (Monogenea: Dactylogyridae) on six Mediterranean and Black Sea species of grey mullets (Mugilidae). The analyses go along the lines of Russo et al. (2017). However, the Mantel tests proposed by these authors to test whether phylogenetically related taxa are more likely to interact with similar partners could not be implemented here. In specialized associations such as often occurs with host and parasites, both closely and distantly related taxa can be equally disimilar in the associations with their partners. So the Mantel tests lack sufficient resolution. In this particular example, Cophylogenetic signal and the interactions between phylogeny and morphology of hosts and parasites will be assessed with the sum  of squared residuals (*m<sup>2</sup>*) produced by PACo (Balbuena et al. 2013), which is inversely proportional to the concordance between the configurations being tested. Since *m<sup>2</sup>* depends on branch length units, the dissimilarity matrices will be normalized prior to analysis.

The data needed to run this example are embedded in an R object. 
```{r load_data}
LigoMulData <- readRDS("ligophorus_example.rds")
```

The analyses require three R libraries, `phytools` (Revell 2012), `paco` (Hutchinson et al. 2017) and plotly (Sievert 2020). 

```{r libraries}
library(phytools)
library(paco)
library(plotly)
```

### Host Parasite Associations 
The tanglegram below depicts the host-parasite associations based on phylogenetic informacion of *Ligophorus* (González-Rodríguez et al. 2016) and grey mullets (Durand et al. 2012). Data on host-parasite associations is from González-Rodríguez et al. (2017).

```{r tanglegram1}
Mul.tree <- LigoMulData$mullets_tree
Lig.tree <- LigoMulData$lig_tree
HS <- LigoMulData$assoc_mat
  # Arrange host-parasite associations in pairs 
HS.lut <- which(HS ==1, arr.ind=TRUE)
linkhs <- cbind(rownames(HS)[HS.lut[,1]], colnames(HS)[HS.lut[,2]])
obj <- cophylo(Mul.tree,Lig.tree, linkhs, rotate = TRUE)
plot.cophylo(obj, link.lwd=1, link.lty=1, fsize=0.5, pts=FALSE,
             link.col = "seagreen")

```

### Cophylogentic Signal
We apply PACo to the normalized matrices of patristic distances of *Ligophorus* spp. and mullets. The significance of the cophylogentic association between hosts and parasites is established based on 1,000 permutations. 
```{r cophylsignal}
pdLig <- LigoMulData$patriticLig/max(LigoMulData$patriticLig)
pdMul <- LigoMulData$patristicMul/max(LigoMulData$patristicMul)
set.seed(2021) # we set a randomization seed to warrant reproducibility.
CS <- prepare_paco_data(pdMul, pdLig, HS)
CS <- add_pcoord(CS, correction = "cailliez")
CS <- suppressWarnings(PACo(CS, nperm = 1000, symmetric = TRUE, 
                             shuffled = TRUE))
CS$gof
```
The analysis returns a *m<sup>2</sup>* of 0.706 and the cophylogenetic association between *Ligophorus* spp. and grey mullets is significant at the 0.05 level (*p* = 0.014).

### Interaction of Host phylogeny with *Ligophorus* Shape
We assess the concordance of host phylogeny with the shape of ventral anchors of *Ligophorus* spp. Ventral anchors are used for attachment to the gills. If significant, one can conclude that host speciation determined anchor shape of *Ligophorus* spp. Procrustes coordinates of ventral anchors of 244 specimens of *Ligophorus* (González-Rodríguez et al. (2017) are used to compute the Procrustes distances between species.
```{r hostPhylXparaShape}
LigProc <- LigoMulData$procrustes_coord
LigProc$species <- as.factor(LigProc$species)
splitXspp <- split(LigProc[, 2:17], LigProc$species)
# Compute Procrustes coordinates of species centroids:
spp_centroids <- t(sapply(splitXspp, colMeans))
# Euclidean distances between centroids, i.e. Procrustes distances
PDLigo <- as.matrix(dist(spp_centroids))
# Normalize Procrustes distances
PDLigo <- PDLigo/max(PDLigo)
# Apply PACo
HI <- prepare_paco_data(pdMul, PDLigo, HS)
HI <- add_pcoord(HI, correction = "cailliez")
HI <- suppressWarnings(PACo(HI, nperm = 1000, symmetric = TRUE,
           shuffled = TRUE))
HI$gof
```
PACo yields a *m<sup>2</sup>* of 0.792. The association between mullet phylogeny and shape of ventral anchors of *Ligophorus* spp. is not significant at the 0.05 level (*p* = 0.067). 
The concordance (or rather the lack of thereof) between host phylogeny and parasite shape can be visualized as a tanglegram:
```{r tanglegram2}
obj <- cophylo(Mul.tree, as.phylo(hclust(dist(spp_centroids))), linkhs,
               rotate = TRUE)
plot.cophylo(obj, link.lwd=1, link.lty=1, fsize=0.5, pts=FALSE,
             link.col = "firebrick")

```

### Interaction of *Ligophorus* Phylogeny with Mullet Shape
The association between phylogeny of *Ligophorus* spp. and host morphometry is assessed based on data on fishbase.org. (Ideally information of gill shape should have been used but these data are unavailable.)
```{r parPhyloXhostShape}
# Handle the morphometric data from fishbase.org to get a distance matrix
MulMorph <- LigoMulData$mullets_morph
# Standardize variables & compute Euclidean distances between mullet spp.
MulMorph <- scale(MulMorph)
dMulMorph <- as.matrix(dist(MulMorph))
# Normalize distance matrix
dMulMorph <- dMulMorph/max(dMulMorph)
# Apply PACo
SI <- prepare_paco_data(dMulMorph, pdLig, HS)
SI <- add_pcoord(SI, correction = "cailliez")
SI <- suppressWarnings(PACo(SI, nperm = 1000, symmetric = TRUE,
           shuffled = TRUE))
SI$gof
```
We find a higher concordance between *Ligophorus* phylogeny and host shape than cophylogenetic signal (*m<sup>2</sup>* = 0.610 vs. *m<sup>2</sup>* = 0.706). This association is highly significant *p* = 0.003 and again can be inspected with a tanglegram:
```{r tanglegram3}
obj <- cophylo(as.phylo(hclust(dist(MulMorph))), Lig.tree, linkhs,
               rotate = TRUE)
plot.cophylo(obj, link.lwd=1, link.lty=1, fsize=0.5, pts=FALSE,
             link.col = "steelblue")

```

### Translation to Cophylospace
The plot below shows the position of the system studied in cophylospace. For compatibility with the framework described in the accompanying paper, the axes represent 1 - *m<sup>2</sup>*.
``` {r 3dplot}
plot_ly(x = 1-HI$gof$ss, y = 1-SI$gof$ss, z= 1-CS$gof$ss, type="scatter3d",
        mode="markers") %>%
  layout(
    scene = list(
      xaxis = list(title = "Symbiont interacion", range = c(0,1)),
      yaxis = list(title = "Host interaction", range = c(0,1)),
      zaxis = list(title = "Cophylogenetic signal", range = c(0,1))
    )
  )
```

### Discussion
Under a classical cophylogenetic analysis, the significant congruence between mullet and *Ligophorus* phylogenies revealed by PACo would indicate that some degree of cospeciation occurred. However, the evidence presented here indicates that morphometric traits of the hosts can explain speciation of *Ligophorus* better than host phylogeny. The position of our system in cophylospace space (Figure I in Box 2, accompanying paper) suggests some degree of assymmetry in which host traits influenced speciation of *Ligophorus* spp. In fact, the arrangement of *Ligophorus* spp. parasitizing *Mugil cephalus* (*L. mediterraneus*, *L. cephali* and *L. chabaudi*), *P. haematocheilus* (*L. pilengas* and *L. llewellyni*) and *C. saliens* (*L. acuminatus* and *L. minimus*) in respective crown groups seems to indicate several intra-host duplication events. As in other monogeneans (see Blasco-Costa et al. 2012), diversity of microhabitats provided by fish gills could have facilitated intrahost lineage duplications in *Ligophorus*. This together with high host specificity would explain why phylogenetically related *Ligophorus* spp. are more likely to interact with the same host species than expected by chance. In any case, cospeciation does not seem to be the major force accounting for speciation of *Ligophorus* in the Mediterranean and Black Seas.
<br>

#### Funding
<font size = 2>
<p>Funds were provided by the Swiss National Science Foundation (SNSF grant 31003A_169211 to I. Blasco-Costa) and Ministry of Science and Innovation of Spain (grant PID2019-104908GB-I00 to J.A. Balbuena). A. Hayward is supported by a Biotechnology and Biological Sciences Research Council (BBSRC) David Phillips Fellowship (BB/N020146/1).</p>
#### References
<font size = 2>
Balbuena, J.A. et al. (2013) PACo: A Novel Procrustes Application to Cophylogenetic Analysis. PLoS ONE 8, e61048. https://doi.org/10.1371/journal.pone.0061048<br>
Blasco-Costa, I. et al. (2012) Molecular phylogeny of species of Ligophorus (Monogenea: Dactylogyridae) and their affinities within the Dactylogyridae. Parasitology International 61, 619–627. https://doi.org/10.1016/j.parint.2012.06.004<br>
Durand, J.-D. et al. (2012) Systematics of the grey mullets (Teleostei: Mugiliformes: Mugilidae): Molecular phylogenetic evidence challenges two centuries of morphology-based taxonomy. Mol Phylogenet Evol 64, 73–92. https://doi.org/10.1016/j.ympev.2012.03.006<br>
Hutchinson, M.C. et al. (2017) paco: implementing Procrustean Approach to Cophylogeny in R. Methods Ecol Evol 8, 932–940. https://doi.org/10.1111/2041-210X.12736<br>
R Core Team (2021) R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.<br>
Revell, L.J. (2011) phytools: an R package for phylogenetic comparative biology (and other things). Methods Ecol Evol 3, 217–223. https://doi.org/10.1111/j.2041-210X.2011.00169.x<br>
Rodríguez-González, A. et al. (2016) Evolutionary modularity and morphological integration in the haptoral anchor structures of Ligophorus spp. (Monogenea: Dactylogyridae). Parasitol Res 115, 3519–3533. https://doi.org/10.1007/s00436-016-5117-z <br>
Rodríguez-González, A. et al. (2017) Evolutionary morphology in shape and size of haptoral anchors in 14 Ligophorus spp. (Monogenea: Dactylogyridae). PLoS ONE 12, e0178367. https://doi.org/10.1371/journal.pone.0178367<br>
Russo, L. et al. (2017) Quantitative evolutionary patterns in bipartite networks: Vicariance, phylogenetic tracking or diffuse co‐evolution? Methods Ecol Evol 9, 761–772. https://doi.org/10.1111/2041-210X.12914<br>
Sievert C. (2020) Interactive web-based data visualization with R, plotly, and shiny. Chapman and Hall/CRC. ISBN 9781138331457<br>
</font>